---
title:  "Homework 1"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Part I. Changing the author field and file name. (5 points)

#### (a) Change the `author:` field on the Rmd document 

#### (b) Rename this file to "HW1_YourFirstInitialYourLastName.Rmd", (e.g. HW1_YDing.Rmd).

<br>


### Part II. Course Related Questions (30 points)
#### Question 1. What is the instructor's name? Please write about her office and office hours this semester.

> Your Answer:

#### Question 2. Please list the acts of academic dishonesty in the syllabus.

> Your Answer:

#### Question 3. What's the policy for late homework submission? When can you request an extension? How many exams will we have?

> Your Answer:

#### Question 4.  Will we have a final exam? What is the evaluation of the final project? What is the assessment of this class? What is the grade range of A?

> Your Answer:

#### Question 5. Is there a second chance to resubmit your assignment or exam after grading? Are there any extra assignments or work to improve your grade?

> Your Answer:

#### I fully understand the rules in the syllabus and will follow them.

> Your Name Here As Signature:

<br>

### Part III. Basic R Syntax (65 points)

#### Question 1.  Hello World! (10 points)

Here's an R code chunk that prints the text 'Hello world!'.

```{r}
print("Hello world!")
```

#### Modify the code chunk below to print your name and your hobbies

```{r}
# Edit me
```

<br>

#### Question 2. Online Store Sales (20 points)

An online store recorded the number of units sold and the price of five products.

```{r}
product <- c("Keyboard", "Mouse", "Webcam", "Headphones", "USB Drive")
units_sold <- c(12, 25, 8, 15, 30)
unit_price <- c(45, 20, 65, 35, 12)
```

#### a) Product Revenue (5 points)

Calculate the revenue generated by each product. Store the results in a vector named `revenue` and display it.

```{r}
# Edit me
```

#### b) Total and Average Revenue (5 points)

Calculate the total revenue and the average revenue per product.

```{r}
# Edit me
```

#### c) High-Revenue Products (5 points)

Display the names of the products that generated more than $500 in revenue.

```{r}
# Edit me
```

#### d) Popular and Affordable Products (5 points)

Display the names of the products for which more than 15 units were sold and the unit price was less than $25.

```{r}
# Edit me
```

#### Question 3. Working with a Data Frame (35 points)

The following data frame contains health information for three individuals. Height is measured in centimeters, and weight is measured in kilograms.

```{r}
BMI <- data.frame(
  gender = c("Male", "Male", "Female"),
  height = c(152, 171.5, 165),
  weight = c(81, 93, 78),
  Age = c(42, 38, 26)
)

print(BMI)
```

#### a) Calculate BMI (15 points)

Calculate the BMI of each individual, add the results as a new column named `BMI`, and display the updated data frame.

$$
\text{BMI}=\frac{\text{weight (kg)}}{[\text{height (m)}]^2}
$$

```{r}
# Edit me
```

#### b) Filter by BMI (10 points)

Display the `gender` and `Age` of individuals whose BMI is greater than 30.

```{r}
# Edit me
```

#### c) Filter Using Two Conditions (10 points)

Display the `gender` of individuals whose BMI is greater than 25 and whose `Age` is less than 30.

```{r}
# Edit me
```