---
title: "Homework 2"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part I. Changing the author field and file name.

#### (a) Change the `author:` field on the Rmd document 

#### (b) Rename this file to "HW2_YourFirstInitialYourLastName.Rmd", (e.g. HW2_YDing.Rmd).

<br>


### Part II. Cars data (45 points)

We'll look at data frame and plotting in much more detail in later classes.  For a previous of what's to come, here's a very basic example.

For this example we'll use a very simple dataset.  The `cars` data comes with the default installation of R.  To see the first few rows of the data, just type `head(cars)`.

```{r}
head(cars)
```

#### (a) Do a summary for `cars`. What the min,max, mean and median values of `speed` and `dist`? (10 points)

```{r}
# Edit me
```

> Type your answer here


#### (b) Calculate the variance and standard deviation of `speed` and `dist` (5 points)

```{r}
# Edit me
```

> Type your answer here


#### (c) Calulate the average and standard deviation of `dist` when `speed`  is greater than 12 and less than 19. (5 points)

```{r}
# Edit me
```

> Type your answer here



#### (d) Produce a histogram of `speed` with **10 bins**. Please illustrate your observation from the output. (5 points)

```{r}
# Edit me
```

> Type your answer here

#### (e) Produce a histogram of `dist` with **20 bins**. Please illustrate your observation from the output. (5 points)

```{r}
# Edit me
```

> Type your answer here

#### (f) Use the `boxplot` function to create a boxplot of `speed`. Please illustrate your observation from the output. (5 points)

```{r}
# Edit me
```

> Type your answer here

#### (g) Use the `boxplot` function to create a boxplot of `dist`. Please illustrate your observation from the output. (5 points)

```{r}
# Edit me
```

> Type your answer here

#### (h) Use `qqnorm` and `qqline` to test if `dist` follow the normal distribution. (5 points)

```{r}
# Edit me
```

> Type your answer here

<br>

### Part III. Random Variables (20 points)

#### Question 1. Classify the following random variables as discrete or continuous (8 points):

X: the number of automobile accidents per year in Illinois.

> Type your answer here

Y: the length of time to play 18 holes of golf.

> Type your answer here

M: the amount of milk produced yearly by a particular cow.

> Type your answer here

N: the number of eggs laid each month by a hen.

> Type your answer here

#### Question 2: Multiple Choices (12 points):

#### i) A discrete random variable

A) Takes on only countable values

B) Takes on any value in an interval

C) Has a probability density function

> Type your answer here

#### ii) A continuous random variable is different from a discrete random variable because:

A) It can only take integer values

B) It is described using a probability mass function (PMF)

C) It can take an infinite number of values within a given range

> Type your answer here

#### iii) If $X$ is a constant random variable, meaning $X=c$ for some constant $c$, what is $Var(X)$? 

A) 0

B) $c^2$

C) $E[X]$

D) $E[X^2]$

> Type your answer here

#### iv) Which of the following statements about expectation is TRUE?

A) $E[X]$ is always equal to the sample mean

B) $E[X]$ can be estimated by the sample mean when the sample size is large
	
C) $E[X]$ is always equal to the variance of X

D) $E[X]$ is always an integer

> Type your answer here


#### v) If a histogram is skewed to the right, what does it indicate? 

A) The mean is greater than the median

B) The median is greater than the mean

C) The data is symmetric

D) The data has no outliers

> Type your answer here

#### vi) Which of the following is true about outliers in a boxplot?

A) Outliers are always errors in the data

B) Outliers are represented by dots or asterisks outside the whiskers
	
C) Outliers are included in the box

D) Outliers are not shown in a boxplot

> Type your answer here

### Part IV. Normal Distributions and Critial Values (35 points)

#### Question 1: A soft-drink machine is regulated so that it discharges an average of 200 milliliters per cup. If the amount of drink is normally distributed with a standard deviation equal to 15 milliliters,

#### (a) what fraction of the cups will contain more than 224 milliliters?  (5 points)

**Hint**: You can use `pnorm`.

```{r}
# Edit me
```

#### (b) what is the probability that a cup contains between 191 and 209 milliliters? (5 points)

```{r}
# Edit me
```

#### (c) below what value do we get the smallest 25% of the drinks? (5 points)

```{r}
# Edit me
```


#### Question 2. Critical values with `qnorm` 

#### (a) Please use `qnorm` to find the critical values $z_{0.1}$. (5 points)

```{r}
# Edit me
```

#### (b) Find the value of $z_0$ such that $90\%$ of the standard normal $Z$ values lie between $-z_0$ and $z_0$; that is, $P(-z_0\leq Z\leq z_0) = .90$. (10 points)

```{r}
# Edit me
```

#### (c) Which value is larger? Why? (5 points)

> Type your answer here


