---
title:  "Homework 1 Solution"
author: "Dr. Yuhan Ding"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Part I. Changing the author field and file name. (5 points)

#### (a) Change the `author:` field on the Rmd document

Change `Your Name` in the YAML header to your own name.

#### (b) Rename this file to "HW1_YourFirstInitialYourLastName.Rmd", (e.g. HW1_YDing.Rmd).

For example, Dr. Yuhan Ding would name the file `HW1_YDing.Rmd`.

<br>


### Part II. Course Related Questions (30 points)

#### Question 1. What is the instructor's name? Please write about her office and office hours this semester.

> **Solution:** The instructor is Dr. Yuhan Ding. Her office is RE 116b. Her office hours are Monday and Wednesday, 2:00--3:00 PM, or by appointment.

#### Question 2. Please list the acts of academic dishonesty in the syllabus.

> **Solution:** Acts of academic dishonesty include plagiarism, unauthorized collaboration, and copying from online resources or AI tools without permission.

#### Question 3. What's the policy for late homework submission? When can you request an extension? How many exams will we have?

> **Solution:** Homework must be submitted on time, and late homework will not be accepted. If an extension is needed, it should be requested before the deadline. There will be one exam: the midterm exam.

#### Question 4. Will we have a final exam? What is the evaluation of the final project? What is the assessment of this class? What is the grade range of A?

> **Solution:** There is no final exam. Instead, there is a group final project worth 40% of the course grade. The project includes a reproducible R Markdown report and a 15-minute presentation followed by 5 minutes of questions. The course assessment is Homework 20%, Midterm Exam 40%, and Final Project 40%. An overall course percentage of 85% or above earns an A.

#### Question 5. Is there a second chance to resubmit your assignment or exam after grading? Are there any extra assignments or work to improve your grade?

> **Solution:** No. Assignments and exams cannot be resubmitted after grading, and there are no extra assignments or extra work offered to improve a grade.

#### I fully understand the rules in the syllabus and will follow them.

> **Solution:** Students should type their own name as their signature.

<br>

### Part III. Basic R Syntax (65 points)

#### Question 1. Hello World! (10 points)

Here's an R code chunk that prints the text 'Hello world!'.

```{r}
print("Hello world!")
```

#### Modify the code chunk below to print your name and your hobbies

```{r}
print("My name is Yuhan Ding.")
print("My hobbies are reading and traveling.")
```

<br>

#### Question 2. Online Store Sales (20 points)

An online store recorded the number of units sold and the price of five products.

```{r}
product <- c("Keyboard", "Mouse", "Webcam", "Headphones", "USB Drive")
units_sold <- c(12, 25, 8, 15, 30)
unit_price <- c(45, 20, 65, 35, 12)
```

#### a) Product Revenue (5 points)

Calculate the revenue generated by each product. Store the results in a vector named `revenue` and display it.

```{r}
revenue <- units_sold * unit_price
revenue
```

#### b) Total and Average Revenue (5 points)

Calculate the total revenue and the average revenue per product.

```{r}
total_revenue <- sum(revenue)
average_revenue <- mean(revenue)

total_revenue
average_revenue
```

#### c) High-Revenue Products (5 points)

Display the names of the products that generated more than $500 in revenue.

```{r}
product[revenue > 500]
```

#### d) Popular and Affordable Products (5 points)

Display the names of the products for which more than 15 units were sold and the unit price was less than $25.

```{r}
product[units_sold > 15 & unit_price < 25]
```

#### Question 3. Working with a Data Frame (35 points)

The following data frame contains health information for three individuals. Height is measured in centimeters, and weight is measured in kilograms.

```{r}
BMI <- data.frame(
  gender = c("Male", "Male", "Female"),
  height = c(152, 171.5, 165),
  weight = c(81, 93, 78),
  Age = c(42, 38, 26)
)

print(BMI)
```

#### a) Calculate BMI (15 points)

Calculate the BMI of each individual, add the results as a new column named `BMI`, and display the updated data frame.

$$
\text{BMI}=\frac{\text{weight (kg)}}{[\text{height (m)}]^2}
$$

```{r}
BMI$BMI <- BMI$weight / (BMI$height / 100)^2
BMI
```

#### b) Filter by BMI (10 points)

Display the `gender` and `Age` of individuals whose BMI is greater than 30.

```{r}
BMI[BMI$BMI > 30, c("gender", "Age")]
```

#### c) Filter Using Two Conditions (10 points)

Display the `gender` of individuals whose BMI is greater than 25 and whose `Age` is less than 30.

```{r}
BMI[BMI$BMI > 25 & BMI$Age < 30, "gender"]
```
