---
title: "Week 12: Subsect Selection"

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.align = 'center'  , out.width="80%"
)
knitr::include_graphics
```


## Reference
1. Chapter 3 of the textbook Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani, 
*An Introduction to Statistical Learning: with Applications in R*. 

2. Chapter 11 of the textbook Chantal D. Larose
and Daniel T. Larose
*Data Science Using Python and R*. 

## Objective

- Know how to use `stepAIC` to do subset selection (Forward/Backward Selection)

## How to choose the important variables?

The most direct approach is called **all subsets** or **best subsets** regression: we compute the least squares fit for all possible subsets and then choose between them based on
some criterion that balances training error with model size.

However, we often can't examine all possible models, since
they are $2^p$ of them; for example when $p = 40$ there are
over a billion models! Instead we need an automated approach that searches through a subset of them. We discuss two commonly use approaches next.

### Forward Selection

* Begin with the **null model** --- a model that contains an
intercept but no predictors.
* Fit $p$ simple linear regressions and add to the null model the variable that results in the lowest RSS.
* Add to that model the variable that results in the lowest RSS amongst all two-variable models.
* Continue until some stopping rule is satisfied, for example when all remaining variables have a **p-value above some threshold**.

**Revisit `Auto` data set**

To run stepwise regression, you first need to install and open the MASS package.

```{r}
library(MASS)
library(ISLR)
data(Auto)
summary(Auto)
i <- sample(2, nrow(Auto), replace=TRUE, prob=c(0.8, 0.2))
AutoTraining <- Auto[i==1,]
AutoTest <- Auto[i==2,]
# Create a null model 
intercept_only <- lm(mpg ~ 1, data=AutoTraining[,1:8])
# Create a full model
all <- lm(mpg~., data=AutoTraining[,1:8])
# perform forward step-wise regression
forward <- stepAIC (intercept_only, direction='forward',scope = formula(all))
# view results of forward stepwise regression
forward$anova
# view final model
summary(forward)
```

Check the `MAE` and `MSE`

```{r}
library(MLmetrics)
ypred_forward <-predict(object = forward, newdata = AutoTest[,1:8])
MAE(y_pred = ypred_forward, y_true = AutoTest$mpg)
MSE(y_pred = ypred_forward, y_true = AutoTest$mpg)
```

### In-Calss Exercise: Forward Selection

* Please construct a forward selection to predict `salary` in the `Salaries` data set and interpret the model.



#### Backward Selection
* Start with **all variables** in the model.
* Remove the variable with the largest p-value --- that is, the variable that is the least statistically significant.
* The new $(p-1)$-variable model is fit, and the variable with the largest p-value is removed.
* Continue until a stopping rule is reached. For instance, we may stop when all remaining variables have a significant
p-value defined by some significance threshold.

Similarly, we use the same function `stepAIC` from `MASS` package

```{r}
backward <- stepAIC (all, direction='backward')
backward$anova
summary(backward)
#Get MAE and MSE
ypred_backward <-predict(object = backward, newdata = AutoTest[,1:8])
MAE(y_pred = ypred_backward, y_true = AutoTest$mpg)
MSE(y_pred = ypred_backward, y_true = AutoTest$mpg)
```



We can also do **both-direction** step wise selection

```{r}
both <- stepAIC (intercept_only, direction='both',scope = formula(all),trace = 0)
both$anova
summary(both)
#Get MAE and MSE
ypred_both <-predict(object = both, newdata = AutoTest[,1:8])
MAE(y_pred = ypred_both, y_true = AutoTest$mpg)
MSE(y_pred = ypred_both, y_true = AutoTest$mpg)
```


### In-Calss Exercise: Backward Selection

* Please construct a backward selection to predict `salary` in the `Salaries` data set and interpret the model.

* Compare these two models. Is there any difference?




