---
title: "Week 6: Hypothesis Testing on Mean(s)"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.align = 'center'  , out.width="65%"
)
knitr::include_graphics
```

## References
1. __Data Science Foundations with Python with zyLabs__, zyBooks, a Wiley brand. http://www.zybooks.com
2. __Probability & Statistics for Engineers & Scientist__, 9th Edition, Ronald E. Walpole,
Raymond H. Myers,
Sharon L. Myers,
Keying Ye, Prentice Hall

## Objectives

- Know the process and result of a hypothesis test for a population parameter.

- Recognize that a hypothesis test may result in an incorrect conclusion, or one of two possible types of error, and each error has consequences.

- Identify the appropriate inference method to answer questions about one mean, and two means.

- State the null and alternative hypotheses for one mean, and two means hypothesis tests.

- Interpret the test statistic and p-value and make a conclusion about the population parameter(s) based on a p-value.

- Know how to use `t.test`.

## Intro to Hypothesis testing

A **hypothesis test** is a method for evaluating a claim, or hypothesis, about a **population parameter** by examining the statistical evidence against the claim based on a sample. The conclusion of a hypothesis test is a decision that the observed data either indicate the claim is plausible or support an alternative explanation. The following steps outline the general process of conducting a hypothesis test.

1. State null and alternative hypotheses about parameters. The **null hypothesis**, $H_0$ , is typically the by-chance or no-effect explanation, and the **alternative hypothesis**,$H_1$, is typically the explanation of an effect, or difference.

2. Calculate a statistic of interest from the sample data that is used to evaluate the null hypothesis.

3. Determine the **p-value**, or likelihood, of obtaining a statistic at least as extreme as the observed statistic when the null hypothesis is true.

4. Draw a conclusion about the null hypothesis based on the statistical evidence provided by the p-value.






## Inference for Mean(s)

### One mean

For a numerical characteristic of a population, typically the **population mean** is of interest. If we know the population variance, theory-based inference methods for a mean are similar to the methods for a proportion, i.e., using the *standard normal distribution*. However, in general, the population variance is unknown. Therefore, the statistics tend to follow the **t-distribution** rather than the standard normal distribution.

*Remark*:
	The $t$-distribution can still be used even if the population distribution is not **normal**, as long as

- the sample size $n$ is **large**.
- the population distribution is not **too-skewed**.

#### Example 1. One mean hypothesis test

Data are collected on a neutral substance (pH
= 7.0). A sample of the measurements were taken with the data as follows:
$$7.07, 7.00, 7.10, 6.97, 7.00, 7.03,7.01,7.01,6.98,7.08$$
Assume the measurements follow an *approximately normal distribution*.

**Parameter**: A neutral substance pH value, $\mu$


**Hypotheses**:

The null hypothesis states that the mean is 7.0. The alternative hypothesis states the mean is different from 7.0.

\begin{align*}
H_0: \mu = 7.0,\\
H_1: \mu \ne 7.0.
\end{align*}

**Observed data**:

A sample of measurements
$$7.07, 7.00, 7.10, 6.97, 7.00, 7.03,7.01,7.01,6.98,7.08$$

We can have 
$$ \bar{x} = 7.025,\qquad s = 0.044$$

**Test statistic**

The standardized $t$ test statistic's calculation assumes the null hypothesis is true. The standard deviation of the sample mean's sampling distribution is estimated by $s/\sqrt{n}$.

$$ t = \frac{\bar{x}-\mu}{s/\sqrt{n}} = \frac{7.025-7}{0.044/\sqrt{10}} \approx 1.7954$$

This follows a t-distribution with $df = n-1 = 9$ under the null hypothesis.

**Critical Region**

For the two-sided hypothesis
\begin{align*}
H_0: \mu = \mu_0,\\
H_1: \mu \ne \mu_0.
\end{align*}

We reject $H_0$ at significance level $\alpha$ when
the computed $t$-statistic
$$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$$
exceeds $t_{\alpha/2, n-1}$ or is less than $-t_{\alpha/2,n-1}$. That is 
$$t> t_{\alpha/2, n-1} \text{ or } t < -t_{\alpha/2, n-1}$$
These regions are called *critical regions*.

Let's consider $\alpha = 0.05,n =10$. We can see the red *critical regions*.

```{r,echo=FALSE}
# Parameters
alpha <- 0.05
df <- 9

# Critical t value for two-tailed test
t_crit <- qt(1 - alpha/2, df)

# Range for x-axis
x <- seq(-4, 4, length = 500)
y <- dt(x, df)

# Plot t-distribution
plot(x, y, type = "l", lwd = 2, col = "blue",
     main = expression(paste("t-Distribution with df = 9, ", alpha, " = 0.05")),
     ylab = "Density", xlab = "t value")

# Shade critical regions
x_crit_left <- seq(-4, -t_crit, length = 100)
x_crit_right <- seq(t_crit, 4, length = 100)

polygon(c(x_crit_left, rev(x_crit_left)), 
        c(dt(x_crit_left, df), rep(0, length(x_crit_left))), 
        col = "red", border = NA)

polygon(c(x_crit_right, rev(x_crit_right)), 
        c(dt(x_crit_right, df), rep(0, length(x_crit_right))), 
        col = "red", border = NA)

# Add vertical lines at critical values
abline(v = c(-t_crit, t_crit), col = "darkred", lwd = 2, lty = 2)

# Add text annotations
text(-t_crit, 0.02, sprintf("t = %.3f", -t_crit), pos = 2, col = "darkred")
text(t_crit, 0.02, sprintf("t = %.3f", t_crit), pos = 4, col = "darkred")

# Add legend
legend("topright", legend = c("t density", "Critical regions (α = 0.05)"),
       col = c("blue", "red"), lwd = 2, bty = "n")
```

We know here the test statistic is **not** in the critical region. We should *not* reject the null hypothesis.

**p-value**

We can also calculate *p-value* $P(|t| > t_{\text{statistics}})$ of this statistic.

```{r}
tstatistic <- 1.7954
2*pt(-tstatistic,df = 9)
```

The p-value is much larger than $0.05$. We should not reject the hypothesis. 

We can also use `t.test` here.
```{r}
#Need to specify mu
phmeter <-c(7.07,7.00,7.10,6.97,7.00,7.03,7.01,7.01, 6.98,7.08)
t.test(phmeter, mu = 7)
```

When p-value is less than or equal to the level of significance $\alpha$, we **reject** the null hypothesis; otherwise we **do not reject** the null hypothesis.

Notice that the sample size of 10 is rather small. An increase in sample size (perhaps another
experiment) may sort things out.

**Note**: How to choose a good sample size is an advanced topic.

#### In-class Exercise: Tests on a Single Mean (Variance Unknown)

Test the hypothesis that the average content of containers of a particular lubricant is 10 liters if the contents of a random sample of 10 containers are 10.2, 9.7, 10.1, 10.3, 10.1, 9.8, 9.9, 10.4, 10.3, and 9.8 liters. Use a 0.01 level of significance and assume that the
distribution of contents is normal.


### Type I and type II errors

The decision from a hypothesis test is to *either reject the null hypothesis or fail to reject the null hypothesis*. The **significance level**, $\alpha$ , of a hypothesis test is how small the p-value must be to conclude the data provide enough statistical evidence to reject the null hypothesis. The decision is to reject the null hypothesis if the p-value is less than or equal to $\alpha$, and fail to reject the null hypothesis if the p-value is greater than $\alpha$.

In reality, either the null hypothesis is true or the alternative hypothesis is true. Thus, the conclusion from a hypothesis test is either correct or incorrect. Ex: Suppose $H_0$ states that the population mean commute time is 25 minutes. If we reject $H_0$ when the true population mean is actually 25 minutes, we make a Type I error. If we fail to reject $H_0$ when the true population mean is not 25 minutes, we make a Type II error.

- A **type I error**  is rejecting the null hypothesis in favor of the alternative when in reality the null hypothesis is true.

- A **type II error** is failing to reject the null hypothesis when in reality the alternative hypothesis is true. 

$$
\begin{array}{c|cc}
\text{Decision V.S. In Reality}  & H_0 \text{ is true}     & H_0 \text{ is false} \\
\hline
\text{Reject } H_0 &  \text{Type I error} & \text{Correct decision}\\
\text{Fail to reject } H_0 &  \text{Correct decision} & \text{Type II error}
\end{array}
$$


#### Checkpoint Questions: Type I and Type II Errors

A new fitness app estimates users' daily calorie burn using wearable-device
data. Before releasing the app, the development team tests whether the mean
absolute error of the calorie estimates is at most 100 calories, so the app
is sufficiently accurate.

They conduct a hypothesis test on the population mean error \(\mu\).

<div class="match-terms" id="type-errors-match">

<p><strong>Match each description with the correct term.</strong></p>

<p>
<strong>Terms:</strong>
Hypotheses; Type I error; Type II error;
Type I error consequence; Type II error consequence
</p>

<div class="match-item" data-answer="hypotheses">
<p>
<strong>a)</strong>
$$H_0: \mu = 100, \qquad H_1: \mu > 100$$
</p>

<select>
<option value="">Select a term</option>
<option value="hypotheses">Hypotheses</option>
<option value="type1">Type I error</option>
<option value="type2">Type II error</option>
<option value="type1-consequence">Type I error consequence</option>
<option value="type2-consequence">Type II error consequence</option>
</select>

<span class="match-feedback"></span>
</div>


<div class="match-item" data-answer="type1-consequence">
<p>
<strong>b)</strong>
Spend additional time and money to improve the app's accuracy
when the app already meets the desired accuracy level.
</p>

<select>
<option value="">Select a term</option>
<option value="hypotheses">Hypotheses</option>
<option value="type1">Type I error</option>
<option value="type2">Type II error</option>
<option value="type1-consequence">Type I error consequence</option>
<option value="type2-consequence">Type II error consequence</option>
</select>

<span class="match-feedback"></span>
</div>


<div class="match-item" data-answer="type2">
<p>
<strong>c)</strong>
Fail to reject $H_0$ when the true mean error is greater
than 100 calories.
</p>

<select>
<option value="">Select a term</option>
<option value="hypotheses">Hypotheses</option>
<option value="type1">Type I error</option>
<option value="type2">Type II error</option>
<option value="type1-consequence">Type I error consequence</option>
<option value="type2-consequence">Type II error consequence</option>
</select>

<span class="match-feedback"></span>
</div>


<div class="match-item" data-answer="type1">
<p>
<strong>d)</strong>
Reject $H_0$ and conclude that the mean error is greater
than 100 calories when the true mean error is 100 calories.
</p>

<select>
<option value="">Select a term</option>
<option value="hypotheses">Hypotheses</option>
<option value="type1">Type I error</option>
<option value="type2">Type II error</option>
<option value="type1-consequence">Type I error consequence</option>
<option value="type2-consequence">Type II error consequence</option>
</select>

<span class="match-feedback"></span>
</div>


<div class="match-item" data-answer="type2-consequence">
<p>
<strong>e)</strong>
Release the app because the test fails to detect that its mean
error is greater than 100 calories, resulting in inaccurate
calorie estimates for users.
</p>

<select>
<option value="">Select a term</option>
<option value="hypotheses">Hypotheses</option>
<option value="type1">Type I error</option>
<option value="type2">Type II error</option>
<option value="type1-consequence">Type I error consequence</option>
<option value="type2-consequence">Type II error consequence</option>
</select>

<span class="match-feedback"></span>
</div>


<br>

<button type="button" id="check-type-errors">
Check Answers
</button>

<button type="button" id="reset-type-errors">
Reset
</button>

</div>


<script>
document.addEventListener("DOMContentLoaded", function () {

  const container = document.getElementById("type-errors-match");

  const checkButton = document.getElementById("check-type-errors");
  const resetButton = document.getElementById("reset-type-errors");

  checkButton.addEventListener("click", function () {

    const items = container.querySelectorAll(".match-item");

    items.forEach(function (item) {

      const correctAnswer = item.dataset.answer;
      const select = item.querySelector("select");
      const feedback = item.querySelector(".match-feedback");

      if (select.value === "") {

        feedback.textContent = " Please select an answer.";
        feedback.style.color = "orange";

      } else if (select.value === correctAnswer) {

        feedback.textContent = " ✓ Correct";
        feedback.style.color = "green";

      } else {

        feedback.textContent = " ✗ Try again";
        feedback.style.color = "red";

      }

    });

  });


  resetButton.addEventListener("click", function () {

    const items = container.querySelectorAll(".match-item");

    items.forEach(function (item) {

      const select = item.querySelector("select");
      const feedback = item.querySelector(".match-feedback");

      select.value = "";
      feedback.textContent = "";

    });

  });

});
</script>


### Two means

If we know the population variances, theory-based inference methods for a mean are similar to the methods for two proportions. However, in general, the population variance is unknown. We should consider the population variances are equal or not. The good thing is no matter the variances are equal or not, the $t$-statistic is the same. The difference is the degrees of freedom. 

#### Example 2. Comparison of two population means

In a study in which
minerals are transferred from the fungus to the trees
and sugars from the trees to the fungus, 20 northern
red oak seedlings exposed to the fungus Pisolithus tinctorus were grown in a greenhouse. All seedlings were
planted in the same type of soil and received the same
amount of sunshine and water. Half received no nitrogen
at planting time, to serve as a control, and the
other half received 368 ppm of nitrogen in the form
NaNO3. The stem weights, in grams, at the end of 140
days were recorded as follows:

No Nitrogen: 
0.32 0.53 0.28 0.37 0.47 0.43 0.36 0.42 0.38 0.43

Nitrogen: 
0.26 0.43 0.47 0.49 0.52 0.75 0.79 0.86 0.62 0.46

**Parameter**:

$\mu_{NON} =$ mean stem weight of the seedlings without nitrogen 

$\mu_{NIT} =$ mean stem weight of the seedlings with nitrogen 

**Hypotheses**:

\begin{align*}
H_0: \mu_{NIT}  = \mu_{NON} ,\\
H_1: \mu_{NIT} \ne \mu_{NON}.
\end{align*}
where the population means indicate mean weights.

**Observed data**:

$$ n_{NIT} = 10, \bar{x}_{NIT} = 0.565, s_{NIT} = 0.1867  \qquad n_{NON} = 10, \bar{x}_{NON} = 0.399, s_{NON} = 0.0728.$$


**Test statistic**


The $t$ test statistic and the p-value can be obtained using statistical software. 

In `R`, we can use `t.test` for two sample t-test. Don't forget to add the condition `var.equal = TURE`. 

```{r}
noNitro<- c(0.32,0.53 ,0.28, 0.37, 0.47, 0.43, 0.36, 0.42,0.38,0.43)
Nitro<- c(0.26,0.43,0.47,0.49,0.52,0.75,0.79,0.86, 0.62,0.46)
t.test(Nitro,noNitro, var.equal = TRUE, conf.level = .95)
```

Here we make an assumption that the variances are equal. Does that make sense?

```{r}
boxplot(Nitro,noNitro)
```

Can we do a mean test with **different variances**? Yes!

```{r}
t.test(Nitro,noNitro, var.equal = FALSE, conf.level = .95)
```

In this example, both methods lead to the same conclusion: we reject $H_0$.

#### In-class Exercise: Tests on Two Means (Unequal Variance) 

A study was conducted by the Department of
Zoology at Virginia Tech to determine if there is a
significant difference in the density of organisms at
two different stations located on Cedar Run, a secondary
stream in the Roanoke River drainage basin.
Sewage from a sewage treatment plant and overflow
from the Federal Mogul Corporation settling pond enter
the stream near its headwaters. The following data
give the density measurements, in number of organisms
per square meter, at the two collecting stations:

Stations 1:
$$5030,13700,10730,11400,860,2200,4250,15040,4980,11910,8130,26850,17660,
22800,1130,1690$$

Stations 2:
$$2800, 4670,6890,7720,7030,7330,2810,1330,3320,1230,2130,2190$$

Is there sufficient evidence, at the 0.05 significance level, to conclude that the mean densities at the two stations are different?
Assume that the observations come from normal populations
with different variances.

