---
title: "Week 3: Special Discrete Random Distributions"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### References
1. __Probability & Statistics for Engineers & Scientist__, 9th Edition, Ronald E. Walpole,
Raymond H. Myers,
Sharon L. Myers,
Keying Ye, Prentice Hall
2. ITM514 Handout Notes Sampling Distribution, Sonja Petrovic

## Objectives:

- Understand the Bernoulli distributions

- Understand the Binomial distributions

- Understand tossing coin experiments

- Know how to use `barplot`



## Special Discrete Distributions

**Bernoulli and Binomial distributions** are related to experiments with binary outcome (success or failure).

#### Bernoulli Distribution

**Bernoulli Trial**: A single random experiment with outcome either success or failure, and the probability of the success is $p$. 

*Examples*
-	Flip a single coin
- Shot a free throw

**Bernoulli Random Variable**: A random variable that takes either $1$ or $0$, where it takes 1 when the outcome of the Bernoulli trial is a success and 0 when the outcome of the Bernoulli trial is a failure.

**Bernoulli Distribution**: The distribution of a Bernoulli random variable is called the Bernoulli distribution.

**Notation**: $X\sim Bernoulli(p)$

**PMF of Bernoulli Distribution/Random Variable**:

$$
\begin{array}{c|cc}
x & 0 & 1 \\
\hline
P(X = x) & 1-p & p  \\
\end{array}
$$

**Expectation and Variance**

$$E(X) = (0)(1-p)+(1)p = p,\qquad var(X) = (0)^2(1-p)+(1)^2p -p^2 = p(1-p)$$

#### Investigate $p$ via Tossing Coin Experiments

Suppose we have a fair coin and consider head as a success. Thus $p = 0.5$. 
We generate 10 instances from the Bernoulli distribution.

```{r, echo=TRUE,fig.width = 6, fig.height = 3}
# number of random experiments
n <- 10
# sample space for tossing a coin, 0--Tail, 1--Head
x <- c(0,1)
# generate n outcomes of the experiments
coinToss <- sample(x, size=n, replace=TRUE)
# display the results
coinToss
```

We can use `bar plot` to show the frequency of each value in a categorical/discrete variable. 


```{r}
barplot(table(coinToss))
```

This is actually a sample from the distribution with sample size 10. Let's investigate the sample mean.

```{r}
phat <- mean(coinToss)
phat
```

**Question:** What happened if we increase the sample size?

Let $n =100$

```{r, echo=TRUE,fig.width = 6, fig.height = 3}
# number of random experiments
n <- 100 
# sample space for tossing a coin, 0--Tail, 1--Head
x <- c(0,1) 
# generate n outcomes of the experiments
coinToss <- sample(x, size=n, replace=TRUE)
# display the results
barplot(table(coinToss))
phat <- mean(coinToss)
phat
```

Let's try $n = 10,000$.

```{r, echo=TRUE,fig.width = 6, fig.height = 3}
# number of random experiments
n <- 10000
# sample space for tossing a coin, 0--Tail, 1--Head
x <- c(0,1) 
# generate n outcomes of the experiments
coinToss <- sample(x, size=n, replace=TRUE)
# display the results
barplot(table(coinToss))
phat <- mean(coinToss)
phat
```


**Question**: What can we learn from the experiment?

**Conclusion**: Given a sample of **independent and identically distributed (IID)** values, the *sample mean converges to the expectation (true/population mean)*.




#### Binomial Distribution

**Binomial Trial**: Repeat Bernoulli trial independently for a certain number ($n$) of times, each
repetition has the same possible outcomes (success or failure), and the probability of each
success ($p$) is consistent for all trials. This experiment is called Binomial trial.

**Binomial Random Variable**: The number of successes $X$ of a binomial trial is called a Binomial random variable.

**Binomial Distribution**: The distribution of a Binomial random variable is called the binomial distribution

*Examples*:

-	Number of heads in 40 coin flips

-	Number of hits in 20 free throws

**Notation**: $X\sim B(n,p)$, $n$ is the number of Bernoulli trials and $p$ is the probability of success in each trial.

**PMF**: $b(x;n,p) = {n\choose x}p^x(1-p)^{n-x}$, $x=0,1,\ldots,n$

**Mathematical Reason**: $f(x) = P(X=x)$, that is, in $n$ trials, we observe $x$ successes. The probability of an outcome with $x$ successes is $p^x(1-p)^{n-x}$ (the trials are independent). The number of such outcomes is: ${n\choose x}$ (choosing $x$ trials for success in $n$ trials).  

**Expectation and Variance**:
$$E(X) = np, \qquad var(X) = np(1-p).$$
Binomial random variable $X\sim B(n,p)$ can be viewed as the sum of $n$ *independent and identical* Bernoulli random variables, i.e.,
$$X = Y_1+\cdots+Y_n,$$
where $X\sim B(n,p)$, $Y_1,\ldots,Y_n\sim Bernoulli(p)$ and $Y_1,\ldots,Y_n$ are independent.
Thus, by the properties of expected value and variance,
\begin{align*}
E(X) & = E(Y_1+\cdots+Y_n) = E(Y_1)+\cdots+E(Y_n) = np,\\
var(X) &= var(Y_1+\cdots+Y_n) = var(Y_1)+\cdots+var(Y_n) = np(1-p).
\end{align*}

### Quick Check: Discrete Random Variables

This short practice is not graded. Select an answer to receive immediate
feedback.

<style>
.quick-check {
  border: 1px solid #d9e2ec;
  border-radius: 7px;
  padding: 0.55rem 0.85rem;
  margin: 0.6rem 0;
  background: #f8fafc;
}
.quick-check p { margin: 0 0 0.3rem; }
.quick-check label {
  display: block;
  margin: 0.12rem 0;
  line-height: 1.35;
  cursor: pointer;
}
.quiz-feedback {
  min-height: 1.1rem;
  margin-top: 0.25rem;
  font-weight: 600;
}
.quiz-correct { color: #18794e; }
.quiz-incorrect { color: #b42318; }
</style>

<div class="quick-check" data-answer="a">
<p><strong>1. A Bernoulli trial has:</strong></p>
<label><input type="radio" name="discrete-q1" value="a" onchange="checkPractice(this)"> Two possible outcomes</label>
<label><input type="radio" name="discrete-q1" value="b" onchange="checkPractice(this)"> Three possible outcomes</label>
<label><input type="radio" name="discrete-q1" value="c" onchange="checkPractice(this)"> Infinitely many possible outcomes</label>
<div class="quiz-feedback" aria-live="polite"></div>
</div>

<div class="quick-check" data-answer="a">
<p><strong>2. A Bernoulli random variable can take the values:</strong></p>
<label><input type="radio" name="discrete-q2" value="a" onchange="checkPractice(this)"> 0 and 1</label>
<label><input type="radio" name="discrete-q2" value="b" onchange="checkPractice(this)"> Any value between 0 and 1</label>
<label><input type="radio" name="discrete-q2" value="c" onchange="checkPractice(this)"> Any non-negative integer</label>
<div class="quiz-feedback" aria-live="polite"></div>
</div>

<div class="quick-check" data-answer="b">
<p><strong>3. In a Bernoulli distribution, p represents:</strong></p>
<label><input type="radio" name="discrete-q3" value="a" onchange="checkPractice(this)"> The number of trials</label>
<label><input type="radio" name="discrete-q3" value="b" onchange="checkPractice(this)"> The probability of success</label>
<label><input type="radio" name="discrete-q3" value="c" onchange="checkPractice(this)"> The number of successes</label>
<div class="quiz-feedback" aria-live="polite"></div>
</div>

<div class="quick-check" data-answer="b">
<p><strong>4. A binomial random variable represents:</strong></p>
<label><input type="radio" name="discrete-q4" value="a" onchange="checkPractice(this)"> The outcome of one trial</label>
<label><input type="radio" name="discrete-q4" value="b" onchange="checkPractice(this)"> The number of successes in a fixed number of trials</label>
<label><input type="radio" name="discrete-q4" value="c" onchange="checkPractice(this)"> The time required to complete an experiment</label>
<div class="quiz-feedback" aria-live="polite"></div>
</div>

<div class="quick-check" data-answer="a">
<p><strong>5. Which is an example of a binomial random variable?</strong></p>
<label><input type="radio" name="discrete-q5" value="a" onchange="checkPractice(this)"> The number of heads in 20 coin flips</label>
<label><input type="radio" name="discrete-q5" value="b" onchange="checkPractice(this)"> The height of a student</label>
<label><input type="radio" name="discrete-q5" value="c" onchange="checkPractice(this)"> The time spent waiting in line</label>
<div class="quiz-feedback" aria-live="polite"></div>
</div>

<script>
function checkPractice(input) {
  const question = input.closest('.quick-check');
  const feedback = question.querySelector('.quiz-feedback');
  const isCorrect = input.value === question.dataset.answer;

  feedback.textContent = isCorrect ? 'Correct!' : 'Try again.';
  feedback.className =
    'quiz-feedback ' +
    (isCorrect ? 'quiz-correct' : 'quiz-incorrect');
}
</script>
