---
title: 'Week 1: Intro to R, RStudio, Markdown'
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '')
```


## Reference

1. Part of this note is sourced from materials created by Prof. Alexandra Chouldechova from CMU, also distributed under the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License. 

2. R Tutorials, https://www.tutorialspoint.com/r/r_data_types.htm

3. Notes of ITMD/ITMS/STAT 514 Spring 2021 by Dr. Sonja Petrovic 


## Objectives 

- Familiar with basic `R` syntax and structure; 
- Know how to run `R` code in RStudio from within a Markdown document. 

## Installing and loading packages

Just like every other programming language you may be familiar with, R's capabilities can be greatly extended by installing additional "packages" and "libraries".

To **install** a package, use the `install.packages()` command.  You'll want to run the following commands to get the necessary packages for today's lab:

```
install.packages("rmdformats")
install.packages("ggplot2")
install.packages("knitr")
```

You only need to install packages once.  Once they're installed, you may use them by **loading** the libraries using the `library()` command.  For today's lab, you'll want to run the following code

```{r}
library(knitr)   # contains kable() function
```


## In-class exercise: Hello world! (Let's do it together)

1. Open **RStudio** on your machine

2. File > New File > R Markdown ...

3. Change `summary(cars)` in the first code block to `print("Hello world!")`

4. Save your Rmd file as `helloworld.Rmd`

5. Click `Knit HTML` to produce an HTML file.


# R basics 

## Data Types in R

You'll encounter different kinds of data types

- **Logical/Boolean** Direct binary values: `TRUE` or `FALSE` in `R`

```{r}
print(class(TRUE))
```

- **Numeric**: a number like $1.87, 999,1e10$

```{r}
print(class(1.87))
print(class(999))
```

- **Integers**: whole numbers (positive, negative or zero)

```{r}
print(class(2L))
```

- **Characters** fixed-length blocks of bits, with special coding;
In general, **strings** = sequences of characters; `R` only use **Characters** to denote both characters and strings.

```{r}
print(class("TRUE"))
```
- **Missing or ill-defined values**: `NA`, `NaN`, etc.

```{r}
is.na(7)
is.na("seven")
```

- **Special Values**: `pi`，`exp()`

```{r}
pi
exp(1)
```


## Assignment operator
Most variables are created with the **assignment operator**, `<-` or `=`  

```{r}
time_factor <- 12
time_factor
time_in_years = 2.5
time_in_years * time_factor
```


The assignment operator also changes values:
```{r}
time_in_months <- time_in_years * time_factor
time_in_months
time_in_months <- 45
time_in_months
```

  
- Using names and variables makes code: easier to design, easier to debug, less prone to bugs, easier to improve,  and easier for others to read

- Avoid "magic constants"; use named variables

- Use descriptive variable names
  - Good: `num_students <- 35`
  - Bad: `ns <- 35 `

  


## Arithmetic Operators 

Command | Description
--------|-------------
`+,-,*,/` | add, subtract, multiply, divide
`^` | raise to the power of
`%%` | remainder after division (ex: `8 %% 3 = 2`)
`( )` | change the order of operations


```{r}
7 + 5 # Addition
7 - 5 # Subtraction
7 * 5 # Multiplication
7 ^ 5 # Exponentiation
```
## 
```{r}
7 / 5 # Division
7 %% 5 # Modulus
7 %/% 5 # Integer division 
```

## Relational Operators

Each element of the first vector is compared with the corresponding element of the second vector. The result of comparison is a Boolean value.

```{r}
7 > 5
7 < 5
7 >= 7
7 <= 5
```


```{r}
7 == 5
7 != 5
```

## Logical operators

Each element of the first vector is compared with the corresponding element of the second vector. The result of comparison is a Boolean value.

- `&`: Element-wise Logical AND operator.

```{r}
(5 > 7) & (6*7 == 42)
```

- `｜`: Element-wise Logical OR operator.

```{r}
(5 > 7) | (6*7 == 42)
```

- `!`: Logical NOT operator.

```{r}
!(5 > 7) 
```

**Question**: Can we use `&&` or  `||`?

# R objects

## Vectors
  

- A **vector** is one object holds elements of one data type as shown above

- `c()` function returns a vector containing all its arguments in order

```{r}
students_name <- c("Jacob", "Isaac", "Sarah", "Noah", "Joseph")
midterm <- c(80, 90, 93, 82, 95)
```

- Typing the variable name at the prompt causes it to display

```{r}
students_name
```

### Indexing
- `vec[1]` is the first element, `vec[4]` is the 4th element of `vec`
```{r}
students_name[4]
```

- `vec[-4]` is a vector containing all but the fourth element
```{r}
students_name[-4]
```

-  vector of indices:
```{r}
students_name[c(2,4)]
```

- vector of negative indices
```{r}
students_name[c(-1,-3)]
```


### Vector arithmetic
  
Operators apply to vectors "pairwise" or "elementwise":

```{r}
final <- c(78, 84, 95, 82, 91) # Final exam scores
midterm # Midterm exam scores
midterm + final # Sum of midterm and final scores
(midterm + final)/2 # Average exam score
course_grades <- 0.4*midterm + 0.6*final # Final course grade
course_grades
```


## In-Class Exercise: Indexing

1. How to use indexing to get the midterm grade of `Noah`?

2. How to get the students name with `midterm grade` above 90?

3. If the weight changes to 30% for midterm and 70% for final, only the students
whose grade greater than or equal to 90 will get an A. Who will get an A in this course?





### Functions on vectors
Command | Description
--------|------------
`sum(vec)` | sums up all the elements of `vec`
`mean(vec)` | mean of `vec`
`median(vec)` | median of `vec`
`min(vec), max(vec)` | the largest or smallest element of `vec`
`sd(vec), var(vec)` | the standard deviation and variance of `vec`
`length(vec)` | the number of elements in `vec`
`pmax(vec1, vec2), pmin(vec1, vec2)` | example: `pmax(quiz1, quiz2)` returns the higher of quiz 1 and quiz 2 for each student
`sort(vec)` | returns the `vec` in sorted order
`order(vec)` | returns the index that sorts the vector `vec`
`unique(vec)` | lists the unique elements of `vec`
`summary(vec)` | gives a five-number summary  
`any(vec), all(vec)` | useful on Boolean vectors

- Example
```{r}
course_grades
mean(course_grades) # mean grade
median(course_grades)
sd(course_grades) # grade standard deviation
sort(course_grades)
max(course_grades) # highest course grade
min(course_grades) # lowest course grade
```

### Pairwise comparisons
Is the final score higher than the midterm score?
```{r}
midterm 
final
final > midterm
```

Boolean operators can be applied elementwise:
```{r}
(final < midterm) & (midterm > 80)
```
**Note:** The logical operator `&&` and `||` considers only the first element of the vectors and give a vector of single element as output.

### More referencing

`which()` returns the `TRUE` indexes of a Boolean vector:
```{r}
course_grades
a_threshold <- 90 # A grade = 90% or higher
course_grades >= a_threshold # vector of booleans
a_students <- which(course_grades >= a_threshold) # Applying which() 
a_students
students_name[a_students] # Names of A students
```
A different way to solve it
```{r}
students_name[course_grades >= a_threshold]
```


## In-class exercise: More Indexing 

Using the `midterm`, `final`, and `course_grades`

1. Find the students whose course grade is above 85.

2. Find the students whose midterm and final grades both are greater than 85.


## Lists
A list is an R-object which can contain many different types of elements inside it like vectors, functions and even another list inside it. This is often used in your own function to return the values.

```{r}
list1 <- list(c(4,5,6),79.2,cos,exp)
print(list1)
```
How to extract information from the list?

```{r}
length(list1) # get the number of elements inside
list1[1] # access the first element
```

## Matrices

A matrix is a two-dimensional rectangular data set. It can be created using a vector input to the matrix function.

```{r}
# Create a matrix.
M = matrix( c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = TRUE)
print(M)
```

Is there any difference if we choose `FALSE` for `byrow`?

```{r}
# Create a different matrix.
M = matrix( c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = FALSE)
print(M)
```

How to access each element?

```{r}
M[1,2] # The first row and the second column
```

## Factors

Factors are the r-objects which are created using a vector. It stores the vector along with the distinct values of the elements in the vector as labels. The labels are always character irrespective of whether it is numeric or character or Boolean etc. in the input vector. They are useful in statistical modeling.

Factors are created using the `factor()` function. The `nlevels` functions gives the count of levels.

```{r}
# Create a vector.
apple_colors <- c('green','green','yellow','red','red','red','green')

# Create a factor object.
factor_apple <- factor(apple_colors)

# Print the factor.
print(factor_apple)
print(nlevels(factor_apple))
```
We can also use `as.factor()` to convert a numeric vector to a factor

```{r}
numbers <- c(1,1,2,3,2,1,2)
print(numbers)
print(class(numbers))
print(as.factor(numbers))
print(class(as.factor(numbers)))
```
**Note**: Some class labels are stored as numeric. However, if you want to do classification, you need to convert it to a factor.

## Data Frames
**Data frames** are _tabular data objects_. Unlike a matrix in data frame each column can contain different modes of data. The first column can be numeric while the second column can be character and third column can be logical. It is a list of vectors of equal length.

Data Frames are created using the `data.frame()` function.

```{r}
BMI <- 	data.frame(
   gender = c("Male", "Male","Female"), 
   height = c(152, 171.5, 165), 
   weight = c(81,93, 78),
   Age = c(42,38,26)
)
print(BMI)
```

How to access each column/attribute? How to access each row/observation?

```{r}
BMI$height # to access height
BMI[,2] # a different way to access height
BMI[2,] # to access the second row
```


## In-class exercise: More Indexing II

1. Please create a data frame called `ITMD514Grades` with `name`,`midterm`, and `final`.

2. Add one column named `final_grade` to the data frame using the weight of 30% for midterm and 70% for final

3. Find the students whose final grade is above 85.

4. Find the students whose midterm and final grades both are greater than 85.


## Full info page on Markdown

[Link to extensive resource](https://learninglab.gitlabpages.inria.fr/mooc-rr/mooc-rr-ressources/module1/ressources/introduction_to_markdown.html) with many examples you can reference as needed. 


