---
title: "Week 7: Inference on Proportion(s)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.align = 'center'  , out.width="65%"
)
knitr::include_graphics
# Load libraries
library(ggplot2)
library(patchwork)
```


## References
1. __Data Science Foundations with Python with zyLabs__, zyBooks, a Wiley brand. http://www.zybooks.com
2. __Probability & Statistics for Engineers & Scientist__, 9th Edition, Ronald E. Walpole,
Raymond H. Myers,
Sharon L. Myers,
Keying Ye, Prentice Hall


## Objectives

- Know how to **use R** to construct a confidence interval(CI) on **Proportion** and **Difference of Two Proportions**

- Identify the appropriate inference method to answer questions about one proportion and two proportions.

- Interpret the test statistic and p-value and make a conclusion about the proportion(s) based on a p-value.

-  Know how to use `prop.test`.

## Interval Estimate of a Proportion/Difference between Proportions

A point estimator of the proportion $p$ in a binomial experiment is given by the
statistic  $\hat{p}  =x/n,$ where $x$ represents the number of successes in $n$ trials. If the unknown proportion $p$ is not expected to be too close to 0 or 1, we can
establish a confidence interval for p by considering the sampling distribution of
$\hat{p}.$ 
The sample proportion $\hat{p} = x/n$ is the sample mean of these $n$ values.
Hence , by the Central Limit Theorem, for $n$ is sufficiently large, $\hat{P}$ is approximately normally distributed with **mean**
$\mu_{\hat{P}} = p$
and **variance**
$\sigma^2_{\hat{p}} = \frac{p(1-p)}{n}.$
Therefore 
$$ Z = \frac{\hat{p}-p}{\sqrt{p(1-p)/n}} \sim \mathcal{N}(0,1)$$
where $z_{\alpha/2}$ is the $z$-value leaving an area of $\alpha/2$ to the right.

*Note:* If we don't have the exact $p$, we will use $\hat{p}$ to approximate $p$ under the radical sign.

### Large-Sample Confidence Interval on $p$

If $\hat{p}$ is proportion of successes in a random sample of size $n,$
an approximate $100(1 -\alpha)\%$ confidence interval, for the binomial parameter $p$
is given by 
$$ \hat{p} - z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}} < p < \hat{p} + z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}},$$
where $z_{\alpha/2}$ is the $z$-value leaving an area of $\alpha/2$ to the right.

**Example 1**
In a random sample of $n = 500$ families owning television sets in the city of Hamilton,
Canada, it is found that 
$x = 340$ subscribe to HBO. Find a 95\% confidence
interval for the actual proportion of families with television sets in this city that
subscribe to HBO.

```{r}
n <- 500 # sample size is large enough
x <- 340
phat <- x/n
sigma <- sqrt((1-phat)*phat)
alpha <- 0.05
loBound <- phat - qnorm(1-alpha/2)*sigma/sqrt(n)
loBound
upBound <- phat + qnorm(1-alpha/2)*sigma/sqrt(n)
upBound
```


Similarly, we can obtain a confidence interval on $p_1-p_2$.

### Check Point Questions:

<div id="hbo-ci-checkpoint">

<div class="checkpoint-item" data-answer="c">
<p>
In a random sample of $n=500$ families owning television sets in the city of Hamilton, Canada, it is found that $x=340$ subscribe to HBO. The 95% confidence interval is $(0.637, 0.720)$ and can be interpreted as follows:
</p>

<p>
The researchers are 95% confident that the proportion of _____ with HBO subscriptions is between 0.637 and 0.720.
</p>

<label><input type="radio" name="hbo-q1" value="a"> A. all families owning television sets in Canada</label><br>
<label><input type="radio" name="hbo-q1" value="b"> B. sampled families owning television sets in Canada</label><br>
<label><input type="radio" name="hbo-q1" value="c"> C. all families owning television sets in the city of Hamilton</label><br>
<label><input type="radio" name="hbo-q1" value="d"> D. sampled families owning television sets in the city of Hamilton</label>

<p class="feedback"></p>

<div class="explanation" style="display:none;">
<strong>Answer: C.</strong>
A confidence interval estimates a <strong>population proportion</strong>, not the proportion in the sample. The population of interest is all families owning television sets in the city of Hamilton.
</div>
</div>

<br>

<button type="button" id="check-hbo-ci">Check Answer</button>
<button type="button" id="reset-hbo-ci">Reset</button>

</div>

<script>
document.addEventListener("DOMContentLoaded", function () {

  const container = document.getElementById("hbo-ci-checkpoint");
  const checkButton = document.getElementById("check-hbo-ci");
  const resetButton = document.getElementById("reset-hbo-ci");

  checkButton.addEventListener("click", function () {

    const item = container.querySelector(".checkpoint-item");
    const correctAnswer = item.dataset.answer;
    const selected = item.querySelector('input[type="radio"]:checked');
    const feedback = item.querySelector(".feedback");
    const explanation = item.querySelector(".explanation");

    if (!selected) {
      feedback.textContent = "Please select an answer.";
      feedback.style.color = "orange";
      explanation.style.display = "none";
    } else if (selected.value === correctAnswer) {
      feedback.textContent = "✓ Correct";
      feedback.style.color = "green";
      explanation.style.display = "block";
    } else {
      feedback.textContent = "✗ Incorrect";
      feedback.style.color = "red";
      explanation.style.display = "block";
    }

  });

  resetButton.addEventListener("click", function () {

    const item = container.querySelector(".checkpoint-item");
    const selected = item.querySelector('input[type="radio"]:checked');

    if (selected) {
      selected.checked = false;
    }

    item.querySelector(".feedback").textContent = "";
    item.querySelector(".explanation").style.display = "none";

  });

});
</script>


### Large-Sample Confidence Interval on $p_1-p_2$


If $\hat{p}_1- \hat{p}_2$ are the proportions of successes in random samples of size $n_1$ and $n_2$, respectively, $\hat{q}_1 = 1-\hat{p}_1$, and $\hat{q}_2 = 1-\hat{p}_2$,
an approximate $100(1 -\alpha)\%$ confidence interval for the difference of two binomial parameters, $p_1-p_2$,
is given by 
$$ (\hat{p}_1-\hat{p}_2) - z_{\alpha/2}\sqrt{\frac{\hat{p}_1\hat{q}_1}{n_1} + \frac{\hat{p}_2\hat{q}_2}{n_2}} < p_1-p_2 < (\hat{p}_1-\hat{p}_2) + z_{\alpha/2}\sqrt{\frac{\hat{p}_1\hat{q}_1}{n_1} + \frac{\hat{p}_2\hat{q}_2}{n_2}} ,$$
where $z_{\alpha/2}$ is the $z$-value leaving an area of $\alpha/2$ to the right.

**Example 2**
	A certain change in a process for manufacturing component parts is being considered.
	Samples are taken under both the existing and the new process so as
	to determine if the new process results in an improvement. If 75 of 1500 items
	from the existing process are found to be defective and 80 of 2000 items from the
	new process are found to be defective, find a 90% confidence interval for the true
	difference in the proportion of defectives between the existing and the new process.
	
```{r}
n1<-1500
n2<-2000
x1<-75
x2<-80
p1hat<- x1/n1
p2hat<- x2/n2
se<- sqrt(p1hat*(1-p1hat)/n1+p2hat*(1-p2hat)/n2)
alpha <- 0.1
loBound<- p1hat-p2hat - qnorm(1-alpha/2)*se
loBound
upBound<- p1hat-p2hat + qnorm(1-alpha/2)*se
upBound
```


**Question**: Why don't we need to think about the variance known or unknown for estimation on proportions?

**Answer**: We can use $\hat{p}$ to estimate the variance.

#### In-class Exercise: Interval Estimate Difference between Proportions

A university wants to compare pass rates between two teaching methods.

Method A: 145 out of 160 students passed

Method B: 138 out of 170 students passed

Please construct a 95% CI on the difference between the pass rates.

  
## Hypothesis Testing on Proportion(s)

### One Proportion

**Revisit Example 1**.In a random sample of $n=500$ families owning television sets in the city of Hamilton, Canada, it is found that $x=340$ subscribe to HBO. 

Suppose we make the conjecture, the proportion of families with television sets in this city that subscribe to HBO is 0.7.

**Parameter**:

The population parameter of interest is the **proportion**, $p$, of  families with television sets in Hamilton, Canada that subscribe to HBO
**Hypotheses**:

\begin{align*}
H_0: p & = 0.7, \\
H_1: p & \ne 0.7. 
\end{align*}

**Observed data**:

In a random sample of $n=500$ families owning television sets in the city of Hamilton, Canada, it is found that $x=340$ subscribe to HBO. 
The observed sample proportion is $\hat{p} = 340/500 \approx 0.68$.

**Test statistic**

A standardized test statistic is typically used. Assuming $n = 500$ is large enough to apply the Central Limit Theorem, the sampling distribution of $\hat{p}$  will be approximately normal. That is

$\hat{p} \sim \mathcal{N}(0.68, 0.7\times(1-0.7)/500) = \mathcal{N}(0.68, 0.02049^2)$

$$ z = \frac{\hat{p}-\mu}{\sigma_{\hat{p}}} = \frac{0.68-0.7}{0.02049} \approx -0.9759$$

**p-value**

$\mathcal{N}(0,1)$ distribution is used to find the p-value, or probability, of obtaining the observed statistic, $z \approx -0.9759$ , or one more extreme in the direction of the alternative hypothesis.

```{r,echo=TRUE}
2* pnorm(-0.9759)
```



In `R`, we can use `prop.test` to do the 1- or 2-sample proportion test

```{r}
# alpha is still 0.05
prop.test(x = 340, n = 500, p = 0.7, alternative = "two.sided", conf.level = 0.95)
```

We can find p-value is much larger than $\alpha = 0.05$, we **fail to reject** $H_0$.

**Note**: `prop.test` in `R` is different from *z-score*. But we get same result.



What about 
\begin{align*}
H_0: p & = 0.7, \\
H_1: p & > 0.7. 
\end{align*}

This is one-sided test. We should calculate the *p-value* with one side.

```{r,echo=TRUE}
1-pnorm(-0.9759)
```

Here *p-value* is much larger than two-sided case. 

We can also use `prop.test` with `alternative ="greater".

```{r,echo=TRUE}
prop.test(x = 340, n = 500, p = 0.7, alternative = "greater", conf.level = 0.95)
```

We also can find p-value is much larger than $\alpha = 0.05$, we **fail to reject** $H_0$.


### Two Proportions

A hypothesis test is also useful for comparing the parameters of two independent populations.

**Example 3**. A vote is to be taken among the *residents of a town and the surrounding county*
to determine whether a proposed chemical plant should be constructed. The construction
site is within the town limits, and for this reason many voters in the
county believe that the proposal will pass because of the large proportion of town
voters who favor the construction. To determine if there is a significant difference
in the proportions of town voters and county voters favoring the proposal, a poll is
taken. If 120 of 200 town voters favor the proposal and 240 of 500 county residents
favor it, would you agree that the proportion of town voters favoring the proposal is
higher than the proportion of county voters? Use an $\alpha = 0.05$ level of significance.


**Parameter**:

$p_T =$ proportion of town voters favor the proposal  

$p_C =$ proportion of  county voters favor the proposal 

A hypothesis test can be used to determine the proportion of town voters favoring the proposal is
higher than the proportion of county voters.

**Hypotheses**:

The null hypothesis can be stated as either the two proportions are equal, or the difference between the two proportions is 0. The alternative hypothesis is greater.

\begin{align*}
H_0: p_T & = p_C \text{ or } p_T -p_C =0, \\
H_1: p_T & > p_C. 
\end{align*}

**Note**: Theory-based inference methods for two population proportions require a *large enough random sample* from each population for the **Central Limit Theorem** to apply and the test statistic to be normally distributed.


**Test statistics**:

Under the null hypothesis $H_0:p_1=p_2$, the expected difference between the two sample proportions is 0. For large samples, the difference in sample proportions is approximately normally distributed. Therefore, we standardize the observed difference to obtain a $z$-test statistic.

$$z = \frac{\hat{p_T}-\hat{p_C} -0}{\sqrt{\frac{\hat{p}_T(1-\hat{p}_T)}{n_T} + \frac{\hat{p}_C(1-\hat{p}_C)}{n_C}}} \sim \mathcal{N}(0,1).$$

Based on the example, we have
$$\hat{p}_T = \frac{120}{200} \approx 0.6, \qquad \hat{p}_C = \frac{240}{500} \approx 0.48, \qquad n_T = 200 \qquad n_C = 500.$$

Thus,
$z \approx 2.86972.$

**p-value**:

$\mathcal{N}(0,1)$ distribution is used to find the p-value, or probability, of obtaining the observed statistic, $z \approx 2.86972$ , or one more extreme in the direction of the alternative hypothesis.

```{r}
1-pnorm(2.86972)
```


Therefore, we reject $H_0$ and agree that the proportion of town voters favoring the proposal is **higher** than the proportion of county voters.

Can we use `prop.test`? Yes

```{r}
# alpha is still 0.05
prop.test(x = c(120,240), n = c(200,500), alternative = "greater", conf.level = 0.95)
```

We also can find p-value is much smaller than $\alpha = 0.05$, we  **reject** $H_0$.


<br>

#### Checkpoint Questions: One-Sided vs. Two-Sided Hypothesis Test for Two Proportions

<div id="one-two-sided-checkpoint">

<div class="checkpoint-item" data-answer="b">
<p>
<strong>1)</strong>
The value of the test statistic for the two-sided hypothesis test is
_____ 2.86972, the value of the test statistic for the one-sided
hypothesis test.
</p>

<label><input type="radio" name="q1" value="a"> a) less than</label><br>
<label><input type="radio" name="q1" value="b"> b) equal to</label><br>
<label><input type="radio" name="q1" value="c"> c) greater than</label>

<p class="feedback"></p>

<div class="explanation" style="display:none;">
<strong>Answer: b) equal to.</strong>
Because the test statistic calculation does not depend on the
alternative hypothesis, the one-sided and two-sided hypothesis tests
have the same $z$ test statistic.
</div>
</div>

<br>

<div class="checkpoint-item" data-answer="c">
<p>
<strong>2)</strong>
The p-value for the two-sided hypothesis test is _____ 0.0021,
the p-value of the one-sided hypothesis test.
</p>

<label><input type="radio" name="q2" value="a"> a) less than</label><br>
<label><input type="radio" name="q2" value="b"> b) equal to</label><br>
<label><input type="radio" name="q2" value="c"> c) greater than</label>

<p class="feedback"></p>

<div class="explanation" style="display:none;">
<strong>Answer: c) greater than.</strong>
The two-sided hypothesis test p-value is the area under the
$\mathcal{N}(0,1)$ distribution curve for values greater than or equal to
$z=2.86972$ and less than or equal to $z=-2.86972$.
The two-sided p-value includes both tails and is twice the
one-sided p-value. Therefore, the two-sided p-value is greater
than the one-sided p-value.
</div>
</div>

<br>

<button type="button" id="check-one-two-sided">Check Answers</button>
<button type="button" id="reset-one-two-sided">Reset</button>

</div>

<script>
document.addEventListener("DOMContentLoaded", function () {

  const container =
    document.getElementById("one-two-sided-checkpoint");

  const checkButton =
    document.getElementById("check-one-two-sided");

  const resetButton =
    document.getElementById("reset-one-two-sided");

  checkButton.addEventListener("click", function () {

    const items =
      container.querySelectorAll(".checkpoint-item");

    items.forEach(function (item) {

      const correctAnswer = item.dataset.answer;

      const selected =
        item.querySelector('input[type="radio"]:checked');

      const feedback =
        item.querySelector(".feedback");

      const explanation =
        item.querySelector(".explanation");

      if (!selected) {

        feedback.textContent = "Please select an answer.";
        feedback.style.color = "orange";
        explanation.style.display = "none";

      } else if (selected.value === correctAnswer) {

        feedback.textContent = "✓ Correct";
        feedback.style.color = "green";
        explanation.style.display = "block";

      } else {

        feedback.textContent = "✗ Incorrect";
        feedback.style.color = "red";
        explanation.style.display = "block";

      }

    });

  });

  resetButton.addEventListener("click", function () {

    const items =
      container.querySelectorAll(".checkpoint-item");

    items.forEach(function (item) {

      const selected =
        item.querySelector('input[type="radio"]:checked');

      if (selected) {
        selected.checked = false;
      }

      item.querySelector(".feedback").textContent = "";
      item.querySelector(".explanation").style.display = "none";

    });

  });

});
</script>



```{r, echo = FALSE}

# Given z statistic
z <- 2.86972

# Create data for the standard normal curve
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)
df <- data.frame(x, y)

# --- One-sided (right-tail) ---
shade_one <- subset(df, x >= z)

p1 <- ggplot(df, aes(x, y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_area(data = shade_one, aes(x, y),
            fill = "red", alpha = 0.4) +
  geom_vline(xintercept = z, color = "red",
             linetype = "dashed", linewidth = 1) +
  annotate("text",
           x = z + 0.2,
           y = 0.05,
           label = paste0("z = ", round(z, 2)),
           color = "red") +
  labs(
    title = "One-Sided p-Value (Right Tail)",
    x = "Z value",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)

# --- Two-sided ---
shade_two <- subset(df, x <= -z)
shade_three <- subset(df, x >= z)

p2 <- ggplot(df, aes(x, y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_area(data = shade_two, aes(x, y),
            fill = "red", alpha = 0.4) +
  geom_area(data = shade_three, aes(x, y),
            fill = "red", alpha = 0.4) +
  geom_vline(xintercept = c(-z, z),
             color = "red",
             linetype = "dashed",
             linewidth = 1) +
  annotate("text",
           x = z + 0.2,
           y = 0.05,
           label = paste0("z = ", round(z, 2)),
           color = "red") +
  annotate("text",
           x = -z - 0.4,
           y = 0.05,
           label = paste0("z = -", round(z, 2)),
           color = "red") +
  labs(
    title = "Two-Sided p-Value",
    x = "Z value",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)

# Combine side by side
p1 + p2
```

#### In-class Exercise: Hypothesis Testing on Difference between Proportions

A university wants to compare two teaching methods.

Method A: 145 out of 160 students passed

Method B: 138 out of 170 students passed

The university claimed that Method A is better than Method B. Given the level of significance $\alpha = 0.05$, can you conduct a hypothesis test to verify the claim? What if $\alpha =0.01$?


#### Statistical vs. practical difference

When a hypothesis test results in a **small p-value** and **the null hypothesis is rejected**, the data provide statistical evidence of the parameter(s) being **different from the null hypothesis claim**. However, a statistical difference does not imply a practical difference. Whether or not the difference is enough to be meaningful in real life should also be considered.
